---
title: "Analysis of ARMA(2,1) process"
subtitle: "TMA4285 Time series models - Exercise 3"
author: Anders Christiansen Sørby, Edvard Hove
#bibliography: mybiblio.bib
output: pdf_document
keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Abstract
Blabla

# Introduction

## The ARMA process
A stationary process \(\{X_t\}\) is an \(ARMA(2,1)\) process if the following relation exists
$$
X_t -\phi_1 X_{t-1} - \phi_2 X_{t-2} = Z_t - \theta_1 Z_{t-1}
$$
where \(Z_t \sim \bf{WN}(0, \sigma^2)\).

# Theory

We need to confirm that a given process is indeed an $ARMA(2,1)$


```{r read, echo=FALSE}
x = read.table("DataEx3G8.txt")
```
# Data analysis
The language R has many built-in functions for analysing timeseries. 

## Our data

Our recieved realization of the process can be seen in the next plot.

```{r TSMplot, echo=FALSE}
plot(seq(1,500),x[,1],type="l")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r lagplot, echo=FALSE}
lag.plot(x[,1],12)
```

```{r arma}
arima=arima(x,order=c(2,0,1))
```
```{r acvf}
acf=acf(x)
lines(seq(0,25),ARMAacf(ar=c(0.7411,-0.1282),ma=0.2789,lag.max=25))
n=seq(0,25)
s=0.1*sin(2*pi*(n-6)/14)
lines(n,s)
```
```{r transform}
p=10
c=2*pi/p
y=asin(c*x[,1])
```


# Discussion

Bla bla


# Conclusion

This was ..

# Appendix

Everything not in the main section

# Refrences

